Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-09 ThesisJP/Background/background.tex
Background. ULtrasound imaging.

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Secondly, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}), which has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline. When GPGPU gained interest it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size. When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order.

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction. This is why GPU threads are said to be light weighted. This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions just by thinking of threads.

As the width of SIMD instructions increase the only algorithms that will benefit is those who can be divided into enough concurrent parts. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by instructions. This is something which is impossible to achieve in real life applications.

The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing. A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation. This is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. To generate the ultrasonic frequencies a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the xz-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the yz-plane as well we get what is known as a volumetric or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction dictated by $\Delta_m[n]$.  If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.

Resolution and sampling.

Sampling rates, and data size.

@@

Small section about heart anatomy.

>>>>

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Secondly, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline. When GPGPU gained interest and the GPU was used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which are maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.

When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not depend on the output from other blocks, hence, it must be possible for the GPU to schedule blocks in any order.

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as unit capable of processing one data element in the SIMD instruction. This is why GPU threads are said to be light weighted. This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Threads, or SIMD instructions, is just how one program for the GPU, there is no other way of doing it.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase the only algorithms that will benefit is those who can be divided into enough concurrent fine granular parts. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by issuing instructions. All of this is something which is impossible to achieve in real life applications. One will always end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation. Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g. vascular imaging). To generate the ultrasonic frequencies a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
\caption{Delay-and-sum array beamforming.}
\label{fig:das}
\end{figure}

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.  Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

When generating a B-mode image we have to insure that the spatial sampling is sufficient. We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of an active imaging system is then found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}In order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
FR = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we then get a frame rate of $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. Number of samples in one A-mode image is then found to be:
\begin{align}
N = \frac{2r}{c}B_wf_c.
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency} is the center frequency of the transmitted pulse. The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
For our cardiac probe we have to process close to $50000$ $M$-long vectors of element data every $1/55$ second.

@@

Small section about heart anatomy (Fig.\,\ref{fig:human_heart}).
